Double White Dwarf Mergers with CASTRO\\ I. Methodology and Code Verification

### 2015-01-01 papers/wdmerger_I/paper.tex
Added sentence on initial model section to introduction.
Added a description of the three major methods we're considering for the gravitational coupling to the hydrodynamics.

This is the first in a series of papers designed to address these outstanding theoretical issues for white dwarf mergers. This work will discuss the verification of our hydrodynamics code for simulating these events. Later efforts will look at the initial conditions of the system, the robustness with which a hotspot is found from which a detonation could occur, and the importance of the initial white dwarf models, which should be more sophisticated than simple carbon-oxygen mixtures and in principle should use results from modern stellar evolution calculations. Section \ref{sec:Numerical Methodology} describes our code and why it can provide useful results compared to other methodologies used for this problem. Section \ref{sec:Tests} discusses a few test problems that we use to demonstrate that our code accurately solves the equations of fluid dynamics. Section \ref{sec:Performance} demonstrates that the software scales well for supercomputer applications. Finally, Section \ref{sec:Conclusions and Discussion} recaps what we have shown and highlights some of the future work we plan to do.

@@

The Euler equations for hydrodynamics (in the absence of source terms) in conservative form are:
\begin{align}
  \frac{\partial \rho}{\partial t} &= -\bm{\nabla} \cdot (\rho \mathbf{u}) \\
  \frac{\partial \rho \mathbf{u}}{\partial t} &= -\bm{\nabla} \cdot (\rho \mathbf{u}\mathbf{u}) - \bm{\nabla}p \\
  \frac{\partial \rho E}{\partial t} &= -\bm{\nabla}\cdot(\rho\mathbf{u}E + p\mathbf{u}).
\end{align}
Here $\rho$ is the mass density, $\mathbf{u}$ is the fluid velocity vector, $p$ is the pressure, and $E = \rho \mathbf{u}^2 / 2 + e$ is the total energy density where $e$ is the internal (thermal) energy density.

@@

>>>>

This is the first in a series of papers designed to address these outstanding theoretical issues for white dwarf mergers. This work will discuss the verification of our hydrodynamics code for simulating these events. Later efforts will look at the initial conditions of the system, the robustness with which a hotspot is found from which a detonation could occur, and the importance of the initial white dwarf models, which should be more sophisticated than simple carbon-oxygen mixtures and in principle should use results from modern stellar evolution calculations. Section \ref{sec:Numerical Methodology} describes our code and why it can provide useful results compared to other methodologies used for this problem. Section \ref{sec:initial_models} describes the method we use for setting up a binary white dwarf simulation. Section \ref{sec:Tests} discusses a few test problems that we use to demonstrate that our code accurately solves the equations of fluid dynamics. Section \ref{sec:Performance} demonstrates that the software scales well for supercomputer applications. Finally, Section \ref{sec:Conclusions and Discussion} recaps what we have shown and highlights some of the future work we plan to do.

@@

The Euler equations for hydrodynamics (in the absence of source terms) in conservative form are:
\begin{align}
  \frac{\partial \rho}{\partial t} &= -\bm{\nabla} \cdot (\rho \mathbf{u}) \label{eq:euler_density}\\
  \frac{\partial \rho \mathbf{u}}{\partial t} &= -\bm{\nabla} \cdot (\rho \mathbf{u}\mathbf{u}) - \bm{\nabla}p \label{eq:euler_momentum}\\
  \frac{\partial \rho E}{\partial t} &= -\bm{\nabla}\cdot(\rho\mathbf{u}E + p\mathbf{u}). \label{eq:euler_energy}
\end{align}
Here $\rho$ is the mass density, $\mathbf{u}$ is the fluid velocity vector, $p$ is the pressure, and $E = \mathbf{u}^2 / 2 + e$ is the total energy density where $e$ is the internal (thermal) energy density.

@@

The effect of gravity on the hydrodynamical evolution is typically incorporated by the use of a source term for the momentum and energy equations. The integral form of the momentum source term is
\begin{equation}
  \int \rho \mathbf{g}\, dV
\end{equation}
and for the energy source term it is
\begin{equation}
  \int \rho \mathbf{u}\cdot\mathbf{g}\, dV \label{eq:cell_center_gravity_source}.
\end{equation}
In most hydrodynamics codes these are discretized as $\rho\, \mathbf{g}\, \Delta V$ and $\rho\, \mathbf{u}\,\cdot\mathbf{g}\, \Delta V$, respectively, where $\Delta V$ is the zone volume and $\rho$, $\mathbf{u}$, and $\mathbf{g}$ are evaluated at the zone center. As observed by \cite{arepo}, the essential problem with this approach is that fluid movements in a grid code are calculated using fluxes at zone boundaries, while the standard gravity source term is an estimate of the flux using the zone center quantities. This can lead to a striking mismatch when the gradient across a zone is large. We have found that this is especially important at the edge of a star, where the large density gradient means that the momentum and energy fluxes can transition sharply even across half of a zone width. Another problem is that in a merger calculation in which two stars are moving on a grid, large quantities of material are being advected, so even a slight error in the calculation for the high-density material can manifest itself as a long-term error in energy conservation. This becomes apparent in Section \ref{sec:kepler}. So while a closer look at this may be unwarranted for situations like a stationary single star central explosion calculation, a detailed discussion is certainly called for in the case of an orbiting binary system where getting the gravity right is absolutely necessary for fully trusting the long-term evolution of the system.

\citet[Chapter 4]{shu:1992} observes that it is possible to describe the source term for the momentum equation by taking the divergence of a gravitational stress tensor,
\begin{equation}
  G_{ij} = -\frac{1}{4\pi G}\left(g_i g_j - \frac{1}{2}|\mathbf{g}|^2\delta_{ij}\right).
\end{equation}
The momentum equations are then written explicitly in conservative form when discretized: just as with the momentum stress tensor, divergences are calculated using discretized fluxes at zone boundaries, and the flux at any zone boundary is added to one cell and subtracted from another, so that the total momentum stays constant to within numerical roundoff error (with the possible exception of domain boundaries). This method works because, from the Poisson equation, $\rho$ can be written as proportional to the divergence of $\mathbf{g}$. We note that in order for the method to be reliable, the expression for all components of $\mathbf{g}$ must be correctly calculated from $\phi$ on zone boundaries so that the total divergence does indeed recover the density. This is straightforward for the component normal to the zone interface, but requires averaging over zones in the perpendicular direction to obtain the transverse components.

Dealing with the energy equation is more difficult. The central challenge is to write down a form of the discretized energy equation that explicitly conserves total energy when coupled to gravity. When gravity is included, the total energy is
\begin{equation}
  \rho E_{\text{tot}} = E_G - \frac{1}{2}\rho\phi = \frac{1}{2}\rho \mathbf{u}^2 + \rho e - \frac{1}{2}\rho\phi,
\end{equation}
where $E_G$ is the gas energy from the pure hydrodynamics equation, and the factor of $1/2$ in the gravitational energy term is necessary for simulations with self-gravity to prevent double-counting of interactions. Recently, three different approaches have been proposed for achieving this. We will briefly examine these and then describe the approach that makes the most sense for us.

\citet{marcello:2012} use the mass continuity equation to rewrite the standard source term from Equation \ref{eq:cell_center_gravity_source}. When this is done, the resulting evolution equation for the total energy is:
\begin{equation}
  \frac{\partial (\rho E_{\text{tot}})}{\partial t} = -\bm{\nabla}\cdot\left(\rho \mathbf{u} E_{\text{tot}} + p\mathbf{u} - \frac{1}{2}\rho\mathbf{u}\phi\right) + \frac{1}{2}\phi\frac{\partial \rho}{\partial t} - \frac{1}{2}\rho\frac{\partial \phi}{\partial t}.
\end{equation}
In practice, we would still actually evolve the total gas energy $E_G$, and so when discretized to second order in time from time level $n$ to time level $n+1$ the gas energy equation would look like
\begin{align}
  (\rho E_G)^{n+1} &= (\rho E_G)^{n} + \left(\frac{1}{2} (\rho \phi)^{n+1} - \frac{1}{2} (\rho \phi)^{n}\right) - \Delta t\, \bm{\nabla}\cdot\left(\rho \mathbf{u} E_{\text{tot}} + p\mathbf{u} - \frac{1}{2}\rho\mathbf{u}\phi\right) \notag \\
   &+ \frac{\Delta t}{2}\left(\phi \frac{\partial \rho}{\partial t}\right)^{n+1/2} - \frac{\Delta t}{2} \left(\rho \frac{\partial \phi}{\partial t}\right)^{n+1/2}.
\end{align}
It is not immediately obvious that this is explicitly conservative for the total energy, due to the last two terms on the right-hand side, but written in discrete form the potential is linked to the density in such a way that there is indeed a perfect cancellation of terms between adjacent cells (see Equation 88 in \citeauthor{marcello:2012}). Doing it this way requires a prescription for calculating $\phi^{n+1/2}$, the potential at the time halfway between the new and old state, as well as $\dot{\phi}^{n+1/2}$, the time derivative of the potential at that time. To second order in time, these can be calculated as $(\phi^{n+1} + \phi^{n})/2$ and $(\phi^{n+1}-\phi^{n})/\Delta t$ respectively. So, in practice what we do is to first solve the Poisson equation using $\rho^n$ at the beginning of a time step; then, evolve the pure hydrodynamics equations without the gravitational source to obtain $\rho^{n+1}$; then, solve the Poisson equation for $\phi^{n+1}$; and then to finally calculate the gravitational source terms and add them in. An alternative approach is to use the predictor-corrector scheme that CASTRO features, where the time level $n$ source term is included in the hydrodynamics update and then left out of the final update. In either case, the time level $n$ calculation of $\mathbf{g}$ is included the PPM reconstruction of the edge states.

\citet{jiang:2013} start from the motivation of extending to the energy equation the flux-based formalism that allows explicit conservation for the momentum equation. They determine that the energy is perfectly conserved if we define a gravitational flux $\mathbf{F_g}$ and take its divergence for the energy equation:
\begin{align}
  \mathbf{F_g} &= \frac{1}{8\pi G}\left(\phi\bm{\nabla}\dot{\phi} - \dot{\phi}\bm{\nabla}\phi\right) + \rho\mathbf{u}\phi \label{eq:jiang_flux}\\
  \frac{\partial (\rho E_G)}{\partial t} &= -\bm{\nabla} \cdot \left(\rho\mathbf{u}E_G + p\mathbf{u} + \mathbf{F_g}\right)
\end{align}
This is analytically equivalent to the approach by \citet{marcello:2012}, which can be seen by taking the divergence of the flux and using the Poisson equation. It would be justified to use this form simply because it is cleaner to write down, but there is a more important benefit to using an explicit flux based formula: in AMR, conservation between levels can only happen if all fluid movements are written down using fluxes, because it is those fluxes that are adjusted at coarse-fine interfaces to ensure conservation. In this form we can simply add the gravitational source term to the hydrodynamical fluxes and use the existing AMR algorithm to guarantee conservation between levels. We note that \citeauthor{jiang:2013} say that this method requires the solution of an additional Poisson equation to obtain the intermediate time gravitational potential, but as we pointed out above, this can be done to second order simply by taking a centered difference using the existing Poisson solves, so no extra Poisson solve is necessary.

This flux-based formalism has a significant drawback for our problem: unlike \citeauthor{jiang:2013}, we use open hydrodynamic boundary conditions, not periodic boundary conditions. Mass, momentum and energy can be lost (or gained) across the domain boundary. This is not a serious source of error in the standard cell-centered discretization, because the source term is proportional to $\rho$ and the density near the domain boundary is generally filled with a very low density ambient medium, so the size of the source term is always small. However, the gravitational flux in Equation \ref{eq:jiang_flux} has terms that are not explicitly proportional to density. Consequently, the flux lost across the domain boundaries can be significant (though this effect diminishes with increasing resolution). A related problem is that at the edge of a star, the gravitational flux can be so large that the gas energy becomes negative after an update. This effect is greatly diminished if the flux is instead proportional to density, as in the standard method. This is not a major problem for ideal or polytropic gases in pure hydrodynamics problems, but in our simulations we want the internal energy (and thus the temperature) to always be reliable so that nuclear reactions occur at the correct rates. This is especially sensitive at the edges of the stars, where the burning is expected to occur in a violent merger scneario.

A possible compromise approach is discussed in \cite{arepo}. We start with the integral formulation of the energy source term (Equation \ref{eq:cell_center_gravity_source}), but instead of discretizing at the cell center, we use Gauss' theorem to evaluate the source term at the zone boundaries (see Equations 91--96 of that paper; we exclude the term related to the motion of the zones since our grids are fixed). This results in a gravitational update for zone $i$ written as
\begin{equation}
  \Delta E^{\text{grav}}_{i} = -\frac{1}{2}\sum_{j} \Delta m_{ij}\, \mathbf{r}_{ij} \cdot \bm{\nabla}\phi_i \approx -\frac{1}{2}\sum_{j} \Delta m_{ij}\left(\phi_i - \phi_j\right),
\end{equation}
where the sum is over all $j$ zones that share a boundary with zone $i$. Here $\mathbf{r}_{ij}$ is the radius vector joining zones $i$ and $j$ and $\Delta m_{ij}$ is the mass exchanged between zones $i$ and $j$ in the hydrodynamics update, and is equal to $\Delta t A_{ij} F_\rho^{ij}$, where $A_{ij}$ is the area of the zone boundary and $F_{\rho}^{ij}$ is the mass flux from the mass continuity equation. It is straightforward to see that the second version is explicitly conservative. The benefit of this form is that the mass exchanged between zones near the domain boundary will always be small in absolute terms since these zones are at very low density. Thus the energy lost through the boundary will be of comparable size to the standard method, but the gravity update should be more accurate since the change in gravitational energy now tracks the explicit motion of mass on the grid.

% Now rewrite this so that it is in flux form
